Ab Initio Effective One-Electron Potentials Revisited: General Theory and Example Applications for Charge-Transfer Energy Calculations

### 2019-07-08 papers/2_oep_technique/paper-r0.tex
Paper revision. Added section about multipole distribution

There are two general cases of basis function partitioning scheme
in which it is possible to define an effective one\hyp{}electron
potential. They are listed in Table~\ref{t:oep-matrix-element-types}.
The first type is a Coulomb\hyp{}like contribution, for which multipole expansion or density fitting can be used.
The second type is an overlap\hyp{}like contribution, which can be approximated via the density fitting. Mainly,
%
{
\renewcommand{\arraystretch}{1.4}
\begin{table}[b]
\caption[Types of matrix elements with OEP operators]
{{\bf Types of matrix elements with OEP operators\footnotemark[1]}
}
\label{t:oep-matrix-element-types}
\begin{ruledtabular}
\begin{tabular}{lcccc}
Matrix element      &&            `Overlap-like'                &&            `Coulomb-like'               \\ 
                    && $\tBraKet{i}{\hat{v}^{{\rm eff},A}}{j} $ && $\tBraKet{j}{\hat{v}^{{\rm eff},A}}{l}$ \\ 
	\cline{1-5}
Partitioning scheme &&            $i\in A, j\in B$              &&               $j,l\in B$                \\
Associated ERI's    &&            $\BraKet{AA}{AB}$             &&               $\BraKet{AA}{BB}$         \\
DF\footnotemark[2]/RI\footnotemark[3] Form    
&& $\sum_{\xi\in A}^{\rm Aux} v^A_{i\xi} S^{AB}_{\xi j} $  
&& $\sum_{\xi\zeta\in A}^{\rm Aux} S^{BA}_{j\xi} v^A_{\xi\zeta} S^{AB}_{\zeta l} $ \\
DMTP\footnotemark[4] Form                     
&& --  &  &   \\
\end{tabular}
\end{ruledtabular}
%
\footnotetext[1]{Test footnotemark.}
\footnotetext[2]{Density Fitting}
\footnotetext[3]{Resolution of Identity}
\footnotetext[4]{Distributed Multipole Expansion}
%
\end{table}
}
%

@@

@@

To get the \emph{ab initio} representation of a overlap\hyp{}like OEP, 
one can use a procedure similar to
the typical density fitting or resolution of identity, both of which are nowadays widely used 
to compute electron-repulsion integrals (ERI's) more efficiently.

@@

An arbitrary one-electron potential of molecule $A$ acting on any state vector 
associated with molecule $A$ can be expanded in an auxiliary space centered 
on $A$ as
%
\begin{equation}
   \hat{v}^{A}\Ket{i} = \sum_{\xi\eta}^{\rm RI} \hat{v}^{A}\Ket{\xi} [{\bf S}^{-1}]_{\xi\eta} \BraKet{\eta}{i}
\end{equation}
%
under the necessary assumption that the auxiliary basis set is nearly complete,
i.e., 
$\sum_{\xi\eta}^{\rm RI} \Ket{\xi}[{\bf S}^{-1}]_{\xi\eta} \Bra{\eta} \cong 1$.
In the equations above, 
the `RI' symbol denotes a certain auxiliary basis set that fulfills the resolution of identity. 
%In a special case when the basis set is orthogonal (e.g., molecular orbitals)
%the above relation simplifies to
%\begin{equation}
%   v\vert i) = \sum_{\xi} v\vert \xi) ( \xi \vert i)
%\end{equation}
Such a general expansion can be obtained by 
utilizing the density fitting
in the nearly\hyp{}complete space,
%
\begin{equation}
 \hat{v}^{A}\Ket{i} = \sum_{\xi}^{\rm RI} V^A_{i\xi} \Ket{\xi} \;.
\end{equation}
%
In the above equation,
the matrix ${\bf V}^A$
is the projection of the state vector $\hat{v}^{A} \Ket{i}$
onto the complete basis $\{ \varphi_\xi \}$.
Let $Z_i[{\bf V}^A]$ be the least\hyp{}squares objective function 
%
\begin{equation} \label{e:z-compl}
 Z_i[{\bf V}^A] = \int \vert \Xi_i({\bf r}) \vert^2 d{\bf r}
\end{equation}
%
with the error density defined by
%
\begin{equation}
 \Xi_i({\bf r}) = v^A({\bf r}) \phi_i({\bf r}) - \sum_\xi^{\rm RI} V^A_{i\xi} \varphi_\xi({\bf r}) \;.
\end{equation}
%
By requiring that
%
\begin{equation} \label{e:z-necessary-requirement}
 \frac{\partial Z_i[{\bf V}^A]}{\partial V^A_{i\mu}} = 0 \text{ for all $\mu$}
\end{equation}
%
one finds the coefficients of the $i$th row of ${\bf V}^A$ to be
%
\begin{equation} \label{e:gdf-coml.v}
% {\bf v}_i = {\bf a}^{(i)} \cdot {\bf S}^{-1}
  V^A_{i\xi} = \sum_\eta^{\rm RI} [{\bf S}^{-1}]_{\xi\eta} a^{(i)}_\eta \;,
\end{equation}
%
where the auxiliary matrices are given by
%
\begin{subequations}
\begin{align}
 a^{(i)}_\eta &= \int \varphi^*_\eta({\bf r}) \hat{v}^A \phi_i({\bf r}) d{\bf r} \;,\\  
 S_{\eta\xi}  &= \int \varphi^*_\eta({\bf r}) \varphi_\xi({\bf r}) d{\bf r} \;.
\end{align}
\end{subequations}
%
%Since matrix elements of an OEP operator in auxiliary space can be computed 
%in the same way as the matrix elements with any other basis function, 
%one can formally write the following identity
%\begin{equation}
% (X \vert v\vert i) = \sum_{\xi\eta} S_{X \xi} [{\bf S}^{-1}]_{\xi\eta} (\eta\vert v\vert i)
%\end{equation}
%where $ X $ is an arbitrary orbital.
%When the other orbital
%does not belong to molecule $A$ but to the (changing) environment, it is 
%straightforward to compute the resulting matrix element, which is simply given as  
%\begin{equation}
%   (j_{\in B} \vert v^A \vert i_{\in A}) = \sum_\xi {S_{j\xi}} {G_{i\xi}}
%\end{equation}
%where $j$ denotes the other (environmental) basis function.
%
The working formula for $a^{(i)}_\eta$ can be found by applying 
potential form from Eq.~\eqref{e:v-eff}
along with the spectral representation of the effective density from Eq.~\eqref{e:d-spectral} 
which finally gives
%
\begin{equation}
 a^{(i)}_\eta = \lambda \sum_{x} W_{\eta i}^{(x)} + 
 \sum_{\alpha\beta} D^A_{\alpha\beta} 
  \BraKet{\alpha\beta}{\eta i} \;.
%  \iint \frac{ \phi_\alpha^*({\bf r}_1)  \phi_\beta({\bf r}_1) \varphi^*_\eta({\bf r}_2) \phi_i({\bf r}_2)}
% {\vert {\bf r}_1 - {\bf r}_2 \vert }
% d{\bf r}_1 d{\bf r}_2
\end{equation}
%

@@

\section{\label{s:5}Results and Discussion}
\subsection{\label{ss.5.1}Pauli-Repulsion Interaction Energy}
\subsection{\label{ss.5.2}Charge-Transfer Interaction Energy}

>>>>

There are two general cases of basis function partitioning scheme
in which it is possible to define an effective one\hyp{}electron
potential. They are listed in Table~\ref{t:oep-matrix-element-types}.
The first type is a Coulomb\hyp{}like contribution, for which multipole expansion or density fitting can be used.
The second type is an overlap\hyp{}like contribution, which can be approximated via the density fitting. 
%
{
\renewcommand{\arraystretch}{1.4}
\begin{table}[b]
\caption[Types of matrix elements with OEP operators]
{{\bf Types of matrix elements with OEP operators\footnotemark[1]}
}
\label{t:oep-matrix-element-types}
\begin{ruledtabular}
\begin{tabular}{lcccc}
Matrix element      &&            `Overlap-like'                &&            `Coulomb-like'               \\ 
                    && $\tBraKet{i}{\hat{v}^{A}_{\rm eff}}{j} $ && $\tBraKet{j}{\hat{v}^{A}_{\rm eff}}{l}$ \\ 
	\cline{1-5}
Partitioning scheme &&            $i\in A, j\in B$              &&               $j,l\in B$                \\
Associated ERI's    &&            $\BraKet{AA}{AB}$             &&               $\BraKet{AA}{BB}$         \\
DF\footnotemark[2]/RI\footnotemark[3] Form    
&& $\sum_{\xi\in A}^{\rm Aux} v^A_{i\xi} S^{AB}_{\xi j} $  
&& $\sum_{\xi\zeta\in A}^{\rm Aux} S^{BA}_{j\xi} v^A_{\xi\zeta} S^{AB}_{\zeta l} $ \\
DMTP\footnotemark[4] Form                     
&& --  &  &  $\rho_{jl}^B \odot \rho_{\rm eff}^A$ \\
\end{tabular}
\end{ruledtabular}
%
\footnotetext[1]{Test footnotemark.}
\footnotetext[2]{Density Fitting}
\footnotetext[3]{Resolution of Identity}
\footnotetext[4]{Distributed Multipole Expansion}
%
\end{table}
}
%

@@

This scheme is most applicable in the case of matrix elements of the type
$
 \tBraKet{j^B}{\hat{v}_{\rm eff}^A}{l^B}
$
because it can be considered as a Coulombic interaction between $\rho^B_{jl}$
and $\rho^A_{\rm eff}$. 
In general, given certain two effective one\hyp{}electron density distributions,
the associated effective Coulombic interaction energy can be estimated from the classical formula
according to
%
\begin{equation}
 \Delta^{(XY)}_{\rm eff} E_{\rm Coul} = \iint \frac{\rho_{\rm eff}^X({\bf r}_1) \rho_{\rm eff}^Y({\bf r}_2)}
 {\vert {\bf r}_1 - {\bf r}_2 \vert} 
d{\bf r}_1 d{\bf r}_2 \;.
\end{equation}
%
The above double integral can be approximated by the
distributed multipole (DMTP)
expansions of the respective densities as follows
%
\begin{multline}
  \Delta^{(XY)}_{\rm eff} E_{\rm Coul} \approx
 \rho_{\rm eff}^X \odot \rho_{\rm eff}^Y \equiv
 \sum_{u\in A} \sum_{w\in B} \Big\{ 
 T^{(0)}_{uw}
 q_{\rm eff}^{(u)}  q_{\rm eff}^{(w)} \\
 - {\bf T}^{(1)}_{uw} \cdot 
   \left[ q_{\rm eff}^{(u)} {\BM\upmu}_{\rm eff}^{(w)} - q_{\rm eff}^{(w)} {\BM\upmu}_{\rm eff}^{(u)} \right]
 - {\bf T}^{(2)}_{uw} : 
  {\BM\upmu}_{\rm eff}^{(u)}  \otimes {\BM\upmu}_{\rm eff}^{(w)} 
 \ldots
 \Big\}
\end{multline}
%
In the above equations, $q_{\rm eff}^{(u)}$ and ${\BM\upmu}^{(u)}_{\rm eff}$ 
are the effective distributed monopole (charge)
and dipole moment, respectively, centered on the $u$th site, 
whereas ${{\bf T}^{(r)}_{uw}}$ are the so called interaction tensors or rank $r$ 
that can be found elsewhere.
The symbol `$\odot$' denotes the sum of all the tensor contractions
performed over the DMTP's of molecule $X$ and $Y$ to yield the associated interaction energy.
The choice of the distribution centres as well as the truncation order of the multipole expansion
is crucial in compromising the accuracy and computational cost of the resulting expressions.
There is many ways in which this can be achieved, e.g., through the distributed multipole analysis (DMA)
of Stone and Anderton, %cite
the cumulative atomic multipole moments of Sokalski and Poirier, %cite
the localised distributed multipole expansion (LMTP) of Etchtebest et al. %cite
or schemes based on fitting to electrostatic potential.

@@

This scheme is most useful in the case of matrix elements of the type
$
 \tBraKet{i^A}{\hat{v}_{\rm eff}^A}{j^B}
$.
To get the \emph{ab initio} representation of such an overlap\hyp{}like matrix element,
one can use a procedure similar to
the typical density fitting or resolution of identity, both of which are nowadays widely used 
to compute electron\hyp{}repulsion integrals (ERI's) more efficiently.

@@

An arbitrary one\hyp{}electron potential of molecule $A$ acting on any state vector 
associated with molecule $A$ can be expanded in an auxiliary space centered 
on $A$ as
%
\begin{equation}
   \hat{v}^{A}\Ket{i} = \sum_{\xi\eta}^{\rm RI} \hat{v}^{A}\Ket{\xi} [{\bf S}^{-1}]_{\xi\eta} \BraKet{\eta}{i}
\end{equation}
%
under the necessary assumption that the auxiliary basis set is nearly complete,
i.e., 
$\sum_{\xi\eta}^{\rm RI} \Ket{\xi}[{\bf S}^{-1}]_{\xi\eta} \Bra{\eta} \cong 1$.
In the equations above, 
the `RI' symbol denotes a certain auxiliary basis set that fulfills the resolution of identity. 
%In a special case when the basis set is orthogonal (e.g., molecular orbitals)
%the above relation simplifies to
%\begin{equation}
%   v\vert i) = \sum_{\xi} v\vert \xi) ( \xi \vert i)
%\end{equation}
Such a general expansion can be obtained by 
utilizing the density fitting
in the nearly\hyp{}complete space,
%
\begin{equation}
 \hat{v}^{A}\Ket{i} = \sum_{\xi}^{\rm RI} V^A_{i\xi} \Ket{\xi} \;.
\end{equation}
%
In the above equation,
the matrix ${\bf V}^A$
is the projection of the state vector $\hat{v}^{A} \Ket{i}$
onto the complete basis $\{ \varphi_\xi \}$.
Let $Z_i[{\bf V}^A]$ be the least\hyp{}squares objective function 
%
\begin{equation} \label{e:z-compl}
 Z_i[{\bf V}^A] = \int \vert \Xi_i({\bf r}) \vert^2 d{\bf r}
\end{equation}
%
with the error density defined by
%
\begin{equation}
 \Xi_i({\bf r}) = v^A({\bf r}) \phi_i({\bf r}) - \sum_\xi^{\rm RI} V^A_{i\xi} \varphi_\xi({\bf r}) \;.
\end{equation}
%
By requiring that
%
\begin{equation} \label{e:z-necessary-requirement}
 \frac{\partial Z_i[{\bf V}^A]}{\partial V^A_{i\mu}} = 0 \text{ for all $\mu$}
\end{equation}
%
one finds the coefficients of the $i$th row of ${\bf V}^A$ to be
%
\begin{equation} \label{e:gdf-coml.v}
% {\bf v}_i = {\bf a}^{(i)} \cdot {\bf S}^{-1}
  V^A_{i\xi} = \sum_\eta^{\rm RI} [{\bf S}^{-1}]_{\xi\eta} a^{(i)}_\eta \;,
\end{equation}
%
where the auxiliary matrices are given by
%
\begin{subequations}
\begin{align}
 a^{(i)}_\eta &= \int \varphi^*_\eta({\bf r}) \hat{v}^A \phi_i({\bf r}) d{\bf r} \;,\\  
 S_{\eta\xi}  &= \int \varphi^*_\eta({\bf r}) \varphi_\xi({\bf r}) d{\bf r} \;.
\end{align}
\end{subequations}
%
%Since matrix elements of an OEP operator in auxiliary space can be computed 
%in the same way as the matrix elements with any other basis function, 
%one can formally write the following identity
%\begin{equation}
% (X \vert v\vert i) = \sum_{\xi\eta} S_{X \xi} [{\bf S}^{-1}]_{\xi\eta} (\eta\vert v\vert i)
%\end{equation}
%where $ X $ is an arbitrary orbital.
%When the other orbital
%does not belong to molecule $A$ but to the (changing) environment, it is 
%straightforward to compute the resulting matrix element, which is simply given as  
%\begin{equation}
%   (j_{\in B} \vert v^A \vert i_{\in A}) = \sum_\xi {S_{j\xi}} {G_{i\xi}}
%\end{equation}
%where $j$ denotes the other (environmental) basis function.
%
The working formula for $a^{(i)}_\eta$ can be found by applying 
potential form from Eq.~\eqref{e:v-eff}
along with the spectral representation of the effective density from Eq.~\eqref{e:d-spectral} 
which finally gives
%
\begin{equation}
 a^{(i)}_\eta = \lambda \sum_{x} W_{\eta i}^{(x)} + 
 \sum_{\alpha\beta} D^A_{\alpha\beta} 
  \BraKet{\alpha\beta}{\eta i} \;.
%  \iint \frac{ \phi_\alpha^*({\bf r}_1)  \phi_\beta({\bf r}_1) \varphi^*_\eta({\bf r}_2) \phi_i({\bf r}_2)}
% {\vert {\bf r}_1 - {\bf r}_2 \vert }
% d{\bf r}_1 d{\bf r}_2
\end{equation}
%

@@

\section{\label{s:5}Results and Discussion}
\subsection{\label{ss.5.1}Pauli-Repulsion Interaction Energy}

At short intermolecular interactions, wavefunction overlap increases
which causes the electrons from either molecules to move apart to minimize the repulsive interactions.
The resulting interaction energy can be estimated based on perturbation theory approaches.
In this section, we demonstrate the use of OEP operators to significanlty reduce the computational
cost of such calculations, that can be used in fragment\hyp{}based modelling of extended molecular
aggregates in terms of force field methods.

As a starting point we consider perturbation theory model of Murrell et al.,
in which repulsive energy is originally given by
%
\begin{equation}
 \Delta^{AB} E^{\rm Rep} = 
 \Delta^{AB} E^{\rm Rep}(\mathcal{O}(S)) + 
 \Delta^{AB} E^{\rm Rep}(\mathcal{O}(S^2))
\end{equation}
%
where the first\hyp{}order term with respect to the intermolecular overlap is
%
\begin{multline}
  \Delta^{AB}  E^{\rm Rep}(\mathcal{O}(S)) = -2\sum_{a\in A} \sum_{b\in B}
               S_{ab} \Big\{
           V^A_{ab} 
         + V^B_{ab} \\
 + \sum_{c\in A} \left[ 2(ab \vert cc) - (ac \vert bc) \right] 
 + \sum_{d\in B} \left[ 2(ab \vert dd) - (ad \vert bd) \right]
                \Big\}
\end{multline}
%
and the second\hyp{}order term is given in Eq.(XXX) in Ref. %cite
In this work, we focus to simplify only the former term because this is usually
the bottleneck of evaluating the Pauli repulsive contributions to
forces in molecular dynamics when using the effective potential method, whereas
very inexpensive approximation to the latter term has already been worked out previously. %cite

As another benchmark model for accuracy, we compare our results with 
the exchange and repulsion energy term 
from the density\hyp{}based scheme of partitioning the interaction energy
(here abbreviated as DDS),
that was developed by Mandado and Hermida-Ramon. 
%In the above equations, the Pauli deformation density 
%%
%\begin{equation}
%   \Delta {\bf D}^{\rm Pauli} \equiv {\bf D}^{oo} - {\bf D}
%\end{equation}
%%
%where \f$ {\bf D}^{oo}\f$ and \f$ {\bf D}\f$ are the density matrix formed from
%mutually orthogonal sets of molecular orbitals within the entire aggregate (formed
%by symmetric orthogonalization of MO's) and the density matrix of the unperturbed
%system (that can be understood as a Hadamard sum \f$ {\bf D} \equiv {\bf D}^A \oplus {\bf D}^B\f$).
% 
%At HF level, the Pauli deformation density matrix is given by
%\f[
%  \Delta {\bf D}^{\rm Pauli} = {\bf C} \left[ {\bf S}^{-1} - {\bf 1} \right] {\bf C}^\dagger
%\f]
%whereas the density matrix constructed from mutually orthogonal orbitals is
%\f[
%   {\bf D}^{oo} = {\bf C} {\bf S}^{-1} {\bf C}^\dagger
%\f]
%In the above equations, \f$ {\bf S} \f$ is the overlap matrix between doubly occupied molecular orbitals
% of the entire aggregate.
It is also mentioned here that the DDS model of exchange-repulsion at HF level of theory
is equivalent to the perturbative method of Hayes and Stone and the two methods
yield the same results, although the exchange and repulsive interaction energies
are slightly different.

\subsubsection{\label{ssdff.sd}Defining OEP's for Repulsive Potential}

\subsection{\label{ss.5.2}Charge-Transfer Interaction Energy}
